---
title: "Accessing SpaMTP Annotation Resources"
author: "Tianyao Lu"
date: "2026-08-14"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Accessing SpaMTP Annotation Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`SpaMTPdb` separates versioned RaMP and metabolite annotation resources from
the SpaMTP software package. This keeps the software package small while
allowing analyses to record and reproduce the exact database snapshot used.

```{r}
library(SpaMTPdb)
resources <- SpaMTPdbResources(version = "latest")
resources[c("resource", "version", "category", "default")]
```

# Loading resources

Following AnnotationHub ingestion, a resource is loaded and cached with:

```{r eval=FALSE}
chem_props <- SpaMTPdbResource("chem_props", version = "3.0.7")
```

Package developers can stage the resources locally and configure:

```{r eval=FALSE}
options(SpaMTPdb.resource_dir = "/path/to/SpaMTPdb-resources/3.0.7")
core <- SpaMTPdbBundle()
```

# Session information

```{r}
sessionInfo()
```
