---
title: "Accessing SpaMTP Annotation Resources"
author: "Tianyao Lu"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Accessing SpaMTP Annotation Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`SpaMTPdb` separates versioned RaMP and metabolite annotation resources from
the SpaMTP software package. This keeps the software package small while
allowing analyses to record and reproduce the exact database snapshot used.

Each resource is an independent serialised object, so an analysis downloads
only the components it needs. Resource versions are immutable: future RaMP
releases are added as new versions rather than replacing the RaMP 3.0.7
snapshot.

```{r setup}
library(SpaMTPdb)
resources <- spaMTPdbResources(version = "latest")
resources[c("resource", "version", "category", "default")]
```

# Where the resources are stored

The processed resource files are published on Zenodo and described by metadata
for `r BiocStyle::Biocpkg("AnnotationHub")`. The manifest records, for every
resource, the download location together with the size and MD5 checksum used
to verify a retrieved file.

The same Zenodo version family can also host native experimental RDS used by
SpaMTPData. Publication is coordinated through SpaMTPdb's maintainer recipes,
but the experimental resources stay in SpaMTPData's registry and ExperimentHub
metadata. Sharing the physical host does not merge the two packages or put
experiments into the annotation database bundle. The original RaMP 3.0.7
snapshot keeps its fixed URL; later collection versions have their own URLs.

```{r resource-locations}
resources[c("resource", "file_name", "bytes", "md5")]
```

`spaMTPdbResource()` resolves a resource in four steps, taking the first that
succeeds:

1. a staged local directory, for development and reproducible database builds;
2. an existing download cache that passes its size and checksum checks;
3. the matching AnnotationHub record;
4. the immutable Zenodo source URL recorded in the manifest.

The final step means the package remains usable before the Hub records are
published, and every file retrieved that way is checked against its recorded
size and MD5 checksum before use.

# Inspecting resources without downloading

Passing `metadata = TRUE` returns the registry row and never touches the
network:

```{r inspect-chemical-resource}
chem_props_metadata <- spaMTPdbResource(
  "chem_props",
  version = "3.0.7",
  metadata = TRUE
)
chem_props_metadata[c("resource", "version", "rows", "columns", "bytes")]
```

The pruned RaMP chemical table and the derived structure features remain
separate. SpaMTP's annotation engine loads and joins the second table by
the chemical table's `iso_smiles` to the structure table's `smiles` column
when structure-aware ionisation scoring is requested:

```{r inspect-structure-resource}
smiles_features_metadata <- spaMTPdbResource(
  "smiles_features",
  version = "3.0.7",
  metadata = TRUE
)
smiles_features_metadata[c("resource", "version", "rows", "columns")]
```

# Retrieving a resource

Omit `metadata = TRUE` to retrieve and cache the R object. The evaluated
example uses an explicitly synthetic local fixture, so package checks need
no network connection. `verify = FALSE` is appropriate only for this deliberate
fixture; it must not be used to call modified data an official RaMP snapshot.

```{r retrieve-metadata-resource}
ramp_db_metadata <- local({
  directory <- tempfile("spamtpdb-demo-")
  dir.create(directory)
  on.exit(unlink(directory, recursive = TRUE))
  saveRDS(list(synthetic = TRUE, label = "tutorial fixture"),
    file.path(directory, "ramp_db_metadata.rds"))
  spaMTPdbResource("ramp_db_metadata", local_dir = directory,
    offline = TRUE, verify = FALSE)
})
str(ramp_db_metadata, max.level = 1)
```

Larger resources are retrieved the same way. The chemical property table backs
the indexed MS1 annotation engine:

```r
chem_props <- spaMTPdbResource("chem_props", version = "3.0.7")
```

Downloads are cached, so repeated calls in an analysis resolve from disk. Use
`cache_dir` to override the default user cache returned by
`tools::R_user_dir()`.

# Retrieving a coherent bundle

`spaMTPdbBundle()` returns several resources from one database version, which
keeps identifiers consistent across an analysis:

```{r bundle-metadata}
bundle <- spaMTPdbBundle(
  resources = c("chem_props", "pathway"),
  version = "3.0.7",
  metadata = TRUE
)
vapply(bundle, function(row) row$title, character(1))
```

# Working from a staged directory

Package developers and reproducible database builds can stage the resources
locally and bypass both the Hub and Zenodo entirely:

```r
options(SpaMTPdb.resource_dir = "/path/to/SpaMTPdb-resources/3.0.7")
core <- spaMTPdbBundle()
```

Local files are verified against the registry by default. Setting
`offline = TRUE` uses local files or a verified cache, and errors when neither
is available rather than attempting a network connection.

# The three-package workflow

`SpaMTPdb` supplies versioned annotation tables, `SpaMTPData` supplies experiment
resources and a small `spaMTPExampleData()` SpatialExperiment, and `SpaMTP`
performs analysis using native Bioconductor containers. These data packages do
not import the software package or Seurat. The executable three-package example
is in SpaMTP's `SpaMTP_Companion_Workflow` vignette.

The submission API uses `spaMTPdbResource()`, `spaMTPdbResources()`,
`spaMTPdbBundle()` and `spaMTPdbVersion()`. Update the companion packages together
with SpaMTP when replacing historical PascalCase calls.

# Recording the version used

`spaMTPdbVersion()` reports the available snapshots, newest first. Recording
this value alongside an analysis makes the annotation step reproducible:

```{r versions}
spaMTPdbVersion()
```

# Session information

```{r session-info}
sessionInfo()
```
