---
title: "Accessing SpaMTP Annotation Resources"
author: "Tianyao Lu"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Accessing SpaMTP Annotation Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`SpaMTPdb` separates versioned RaMP and metabolite annotation resources from
the SpaMTP software package. This keeps the software package small while
allowing analyses to record and reproduce the exact database snapshot used.

Each resource is an independent serialised object, so an analysis downloads
only the components it needs. Resource versions are immutable: future RaMP
releases are added as new versions rather than replacing the RaMP 3.0.7
snapshot.

```{r setup}
library(SpaMTPdb)
resources <- SpaMTPdbResources(version = "latest")
resources[c("resource", "version", "category", "default")]
```

# Where the resources are stored

The processed resource files are published on Zenodo and registered with
`r BiocStyle::Biocpkg("AnnotationHub")`. The manifest records, for every
resource, the download location together with the size and MD5 checksum used
to verify a retrieved file.

```{r resource-locations}
resources[c("resource", "file_name", "bytes", "md5")]
```

`SpaMTPdbResource()` resolves a resource in three steps, taking the first that
succeeds:

1. a staged local directory, for development and reproducible database builds;
2. the matching AnnotationHub record;
3. the immutable Zenodo source URL recorded in the manifest.

The third step means the package remains usable before the Hub records are
published, and every file retrieved that way is checked against its recorded
size and MD5 checksum before use.

# Inspecting resources without downloading

Passing `metadata = TRUE` returns the registry row and never touches the
network:

```{r inspect-chemical-resource}
chem_props_metadata <- SpaMTPdbResource(
  "chem_props",
  version = "3.0.7",
  metadata = TRUE
)
chem_props_metadata[c("resource", "version", "rows", "columns", "bytes")]
```

The pruned RaMP chemical table and the derived structure features remain
separate. SpaMTP's annotation engine loads and joins the second table by
`iso_smiles` only when structure-aware ionisation scoring is requested:

```{r inspect-structure-resource}
smiles_features_metadata <- SpaMTPdbResource(
  "smiles_features",
  version = "3.0.7",
  metadata = TRUE
)
smiles_features_metadata[c("resource", "version", "rows", "columns")]
```

# Retrieving a resource

Omit `metadata = TRUE` to retrieve and cache the R object. The provenance
resource is small, so it is a convenient illustration of a real retrieval:

```{r retrieve-metadata-resource}
ramp_db_metadata <- try(
  SpaMTPdbResource("ramp_db_metadata", version = "3.0.7"),
  silent = TRUE
)
if (inherits(ramp_db_metadata, "try-error")) {
  message("Resource retrieval skipped: no network access available.")
} else {
  str(ramp_db_metadata, max.level = 1)
}
```

Larger resources are retrieved the same way. The chemical property table backs
the indexed MS1 annotation engine:

```r
chem_props <- SpaMTPdbResource("chem_props", version = "3.0.7")
```

Downloads are cached, so repeated calls in an analysis resolve from disk. Use
`cache_dir` to override the default user cache returned by
`tools::R_user_dir()`.

# Retrieving a coherent bundle

`SpaMTPdbBundle()` returns several resources from one database version, which
keeps identifiers consistent across an analysis:

```{r bundle-metadata}
bundle <- SpaMTPdbBundle(
  resources = c("chem_props", "pathway"),
  version = "3.0.7",
  metadata = TRUE
)
vapply(bundle, function(row) row$title, character(1))
```

# Working from a staged directory

Package developers and reproducible database builds can stage the resources
locally and bypass both the Hub and Zenodo entirely:

```r
options(SpaMTPdb.resource_dir = "/path/to/SpaMTPdb-resources/3.0.7")
core <- SpaMTPdbBundle()
```

Setting `offline = TRUE` makes a missing local file an error rather than a
download, which is useful on machines without outbound network access.

# Recording the version used

`SpaMTPdbVersion()` reports the available snapshots, newest first. Recording
this value alongside an analysis makes the annotation step reproducible:

```{r versions}
SpaMTPdbVersion()
```

# Session information

```{r session-info}
sessionInfo()
```
