---
title: "Accessing SpaMTP Annotation Resources"
author: "Tianyao Lu"
date: "2026-08-14"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Accessing SpaMTP Annotation Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`SpaMTPdb` separates versioned RaMP and metabolite annotation resources from
the SpaMTP software package. This keeps the software package small while
allowing analyses to record and reproduce the exact database snapshot used.

```{r setup}
library(SpaMTPdb)
resources <- SpaMTPdbResources(version = "latest")
resources[c("resource", "version", "category", "default")]
```

# Loading resources

Inspect a resource without downloading it:

```{r inspect-chemical-resource}
chem_props_metadata <- SpaMTPdbResource(
  "chem_props",
  version = "3.0.7",
  metadata = TRUE
)
chem_props_metadata[c("resource", "version", "rows", "columns")]
```

Following AnnotationHub ingestion, omit `metadata = TRUE` to retrieve and cache
the R object:

```r
chem_props <- SpaMTPdbResource("chem_props", version = "3.0.7")
```

The pruned RaMP chemical table and the derived structure features remain
separate. SpaMTP's annotation engine loads and joins the second table by
`iso_smiles` only when structure-aware ionisation scoring is requested:

```{r inspect-structure-resource}
smiles_features_metadata <- SpaMTPdbResource(
  "smiles_features",
  version = "3.0.7",
  metadata = TRUE
)
smiles_features_metadata[c("resource", "version", "rows", "columns")]
```

Package developers can stage the resources locally and configure:

```r
options(SpaMTPdb.resource_dir = "/path/to/SpaMTPdb-resources/3.0.7")
core <- SpaMTPdbBundle()
```

# Session information

```{r session-info}
sessionInfo()
```
